Gammapy: A Python package for gamma-ray astronomy

### 2023-01-26 src/ms-review.tex
Fix acknowledgements
Signed-off-by: Bruno Khélifi <khelifi@in2p3.fr>
Update src/ms-review.tex
Co-authored-by: Atreyee Sinha <asinha@ucm.es>
author review before publication, reached Sect. 3.5

\newcommand{\hess}{H.E.S.S.\xspace}
\newcommand{\hawc}{HAWC\xspace}
\newcommand{\veritas}{VERITAS\xspace}
\newcommand{\magic}{MAGIC\xspace}
\newcommand{\astri}{ASTRI\xspace}
\newcommand{\iact}{IACT\xspace}
\newcommand{\iacts}{IACTs\xspace}
\newcommand{\cta}{CTA\xspace}
\newcommand{\swgo}{SWGO\xspace}
\newcommand{\irf}{IRF\xspace}
\newcommand{\irfs}{IRFs\xspace}
\newcommand{\fermi}{\textit{Fermi}-LAT\xspace}
\newcommand{\gammaray}{$\gamma$-ray\xspace}
\newcommand{\gammarays}{$\gamma$-rays\xspace}
\newcommand{\gadf}{GADF\xspace}
\newcommand{\milagro}{MILAGRO\xspace}
\newcommand{\github}{GitHub\xspace}

@@

\abstract
	{
		Traditionally, the \gammaray universe has been probed
		by small collaborations of scientists employing proprietary data and analysis software.
		However, the next generation of \gammaray instruments,
		such as the the Cherenkov Telescope Array (CTA), will be operated as open observatories.
		Alongside the data, the observatories will make available to the astronomical community
		software tools for their analysis. This necessity prompted the development, in the last decade,
		of open high-level astronomy software customised for high energy astrophysics.
	}
	{
		In this article, we present \gammapy, an open-source python package for the analysis of \gammaray data,
		and illustrate the functionalities of the first long-term release of the software, version 1.0.
		Built on the modern python scientific ecosystem, \gammapy provides a uniform platform for reducing and
		modelling data from different gamma-ray instruments. \gammapy complies with several well-established
		data conventions in high-energy astrophysics, providing serialised data products that are interoperable
		with other software.
	}
	{
		Starting from event list and "instrument response functions",
		\gammapy provides pipelines for creating reduced data binned in energy and coordinates.
		To facilitate the rejection of the residual hadronic background, several techniques for background estimation
		are implemented in the package.
		After the data are binned, the flux and morphology of one or more \gammaray sources can be estimated
		using Poisson maximum likelihood fitting
		and assuming a variety of spectral, temporal, and spatial models.
		Estimation of flux points, likelihood profiles and light curves is also implemented.
	}
	{
		After describing the structure of the package, we demonstrate the capabilities of \gammapy
		performing both traditional and novel analyses in gamma-ray astronomy with publicly available data:
		spectral and spectro-morphological analyses, estimation of the light curve of a source.
		The flexibility and power of the software are displayed in a final multi-instrument example,
		where data sets from different instruments, at different stages of data reduction,
		are simultaneously fitted with a flux model produced by another astrophysical software.
	}{}

@@

%% \gammaray astronomy
\gammaray astronomy is a rather young field of research. The \gammaray~
range of the electromagnetic spectrum provides us with insight into the
most energetic processes in the universe such as surroundings of
black holes and remnants of supernova explosions. As in other
branches of astronomy, \gammarays can be observed by both
satellite as well as ground based instruments.
Ground based instrument use the Earth's atmosphere as a particle detector.
Very high-energy cosmic \gammarays interact in the atmosphere and
create a large shower of secondary particles that can be observed from the ground.
Ground-based \gammaray astronomy relies on these extensive air showers to detect the
primary \gammaray photons and measure their direction and energy.
It covers the energy range from fews tens of GeV up to the PeV.
There are two main categories of ground based instruments: 
Imaging Atmospheric Cherenkov Telescopes (IACT) and 
Water Cherenkov observatories. IACTs make images of atmospheric showers
by detecting the Cherenkov radiation emitted by the cascading charged particles and
use these images to reconstruct the properties of the incident particle.
Those instruments have a limited field of view (FoV) and duty cycle, but
good energy and angular resolution \citep{2015CRPhy..16..610D}.

\begin{figure*}[t]
	\centering
	\includegraphics[width=\textwidth]{figures/big-picture.pdf}
	\caption{
		Core idea and relation of \gammapy to different \gammaray instruments
		and the gamma astro data formats (GADF). The top left shows the
		group of current and future pointing instruments based on the 
		imaging atmospheric Cherenkov technique (IACT). This includes
		instruments such as the  Cherenkov Telescope Array (CTA),
		the High Energy Stereoscopic System (H.E.S.S.), the
		Major Atmospheric Gamma Imaging Cherenkov Telescopes (MAGIC),
		and the Very Energetic Radiation Imaging Telescope Array System (VERITAS).
		The lower left shows the group of allsky instruments such as the
		Fermi Large Area Telescope (Fermi-LAT) amd the High Altitude 
		Water Cherencov Observatory (HAWC). The calibrated data of all those
		instruments can be converted and stored into the common GADF data format.
		\gammapy can read data stored in the GADF format.
		The \gammapy package is not a part of any instrument, but instead
		provides a common interface to the data and analysis of all
		these \gammaray instruments. This way users can also easily combine data from
		different instruments and perform a joint analysis.
		\gammapy is built on the scientific python eco-system, and the required dependencies
		are shown below the \gammapy logo.
	}
	\label{fig:big_picture}

@@

Water Cherenkov observatories detect particles directly from the tail of the
shower when it reaches the ground. These instruments have very
large field-of-view, large duty-cycle but higher energy threshold and
usually have lower signal to noise ratios compared to IACTs.

%% Context
Ground based \gammaray astronomy has been historically structured
by experiments run by independent collaborations relying
on their own proprietary data and analysis software developed as part of the
instrument. While this model has been successful so far, it does not
permit easy combination of data from several instruments and therefore,
limits the interoperability of existing facilities. This lack of
interoperability currently limits the full exploitation of the
available \gammaray data, especially because the different instruments often have
complementary sky coverages, and the various detection
techniques have complementary properties in terms of covered energy rage,
duty cycle and spatial resolution.

%TODO: better transition into context
The Cherenkov Telescope Array (CTA) will be the first instrument to be operated
as an open observatory in the domain. Its high level data (e.g. the event list) will be shared publicly after
some proprietary period, and the software required to analyze it will be distributed
as well. To allow the re-usability of existing instruments and their interoperability
requires open data formats and open tools that can support the various analysis methods
commonly used in the field.

%% Context : formats
In practice, the data reduction workflow of all \gammaray observatories
is remarkably similar. After data calibration, shower events are reconstructed and
gamma/hadron separation is applied to build lists of \gammaray like events.
The lists of \gammaray events are then used to derive scientific results, such as spectra, sky maps
or light curves, taking into account the observatory specific instrument response functions (IRF).
Once the data is present in the form of a list of events, the information is independent of
the data reduction process, and eventually of the detection technique. This implies,
for example, that data from IACT and particle samplers can be represented
within the same model. The efforts to prototype a format usable by various instruments
converged in the so-called \textit{Data Format for \gammaray Astronomy}
initiative~\citep{gadf_proc, gadf_universe}, abbreviated in
\texttt{gamma-astro-data-format} (\gadf). This proposes prototypical
specifications to produce files based on the flexible image transport system
(FITS) format~\citep{fits} encapsulating this high-level information. This is
realized by storing a list of gamma-like events with their measured quantities
(energy, direction, arrival time) and a parametrisation of the response of the
system.

%% Context: develoment of open software
Python has become extremely popular as a scientific programming language,
in particular, in the field of data sciences. The success is
mostly attributed to the simple and easy to learn syntax, the ability to act as
a 'glue' language between different programming languages and last but not least
the rich eco-system of packages and its open and supportive community, from
core numerical analysis packages such as \numpy~\citep{numpy} and
\scipy~\citep{2020SciPy-NMeth}, visualization libraries, e.g. Matplotlib \citep{matplotlib},
or optimization packages such as \iminuit \citep{iminuit}.

The \astropy~\citep{astropy} was created in 2012 to build a community-developed
Python package for astronomical research. Among other things, it offers functionalities for
transforming and representing astronomical coordinates, manipulating physical quantities
and their units as well as reading and writing FITS files.

%% \gammapy: concept and goals
%%% Question: starting date? Is is relevant, what is the real start?
The \gammapy project started in 2015 with the objective of building a flexible and
efficient software library for very high-energy \gammaray data analysis \citep{gammapy_2015}.
This is possible thanks to the definition of a common data format such as \gadf, and thanks to
an open development with a broad community of contributors. The relation of \gammapy to the
different instruments and the \gadf data format is illustrated in Figure~\ref{fig:big_picture}.

@@

%% \gammapy: from first developments to validation and selection
The \hess collaboration released a limited test dataset (about 50 hours of
observations taken between 2004 and 2008) based  on the \gadf DL3 format \citep{HESS_DR1}.
This data release served as a basis for validation of open analysis tools,
including \gammapy \cite[see e.g.]{Mohrmann2019}.

%% Paper outline
In this article, we describe the general structure of the \gammapy package,
its main concepts and organisational structure. We start in
Section~\ref{sec:gammaray-data-analysis} with a general overview
of the data analysis workflow in very high-energy \gammaray astronomy. Then we
show how this workflow is reflected in the structure of the \gammapy package 
in Section~\ref{sec:gammapy-package} and also
describe the various subpackages it contains. Section~\ref{sec:applications}
presents a number of applications, while Section~\ref{sec:gammapy-project}
finally discusses the project organization.

\section{Gamma-ray Data Analysis}
%
\begin{figure*}[h!]
	\centering
	\includegraphics[width=1.\textwidth]{figures/data_flow.pdf}
	\caption{
		\gammapy sub-package structure and data analysis workflow. The top row
        defines the groups for the different data levels and reduction steps
        from raw gamma-like events on the left, to high level science products
        on the right. The direction of the data flow is illustrated with the
        grey arrows. The gray folder icons represent the different sub-packages
        in \gammapy and their names. Below each icon there is a list of the most
        important objects defined in the sub-package.
    }
	\label{fig:data_flow}
\end{figure*}
%
\label{sec:gammaray-data-analysis}
% This might not be the bast place compared to introduction
The data reduction process in \gammaray astronomy is usually split into two parts.
The first one deals with the data processing from camera measurement, calibration, event
reconstruction and selection to yield a list of reconstructed \gammaray event candidates.
This sequence, sometimes referred to as low-level analysis, is usually very specific to
a given observation technique and even to a given instrument.

The other sequence, referred to as high-level analysis, deals with the extraction of physical
quantities related to \gammaray sources and the production of high-level products such as spectra,
light curves and catalogs. The methods applied here are more generic and are broadly
shared across the field. They also frequently imply joint analysis of multi-instrument data.

@@

To extract physically relevant information, such as the flux, spatial or spectral shape of one or more sources,
an analytical model is commonly adopted to describe the intensity of gamma-ray sources as a function of the energy,
$E_{true}$, and of the position in the field of view, $p_{true}$:
$$ \Phi(p_{\rm true}, E_{\rm true}, \hat{\theta}) \quad \mathrm{[TeV^{-1}cm^{-2}s^{-1}]} $$
where $\hat{\theta}$ is a set of model parameters that can be adjusted in a fit. To convert this analytical flux model
into a prediction on the number of gamma-ray events, $N_{pred}$, with their estimated energy $E$  and position $p$, the model
is convolved through the response function of the instrument.

In the most general way we can write the expected number of detected events from the sky model $\Phi$ at measured position $p$ and energy $E$, for
a given set of parameters $\hat{\theta}$:
\begin{align}
   N(p, E, \hat{\theta}) {\rm d}p {\rm d}E = &t_{\rm obs} \int_{E_{\rm true}} \int_{p_{\rm true}}  R(p, E|p_{\rm true}, E_{\rm true})\\
   &\times \Phi(p_{\rm true}, E_{\rm true}, \hat{\theta} )) {\rm d}E_{\rm true} {\rm d}p_{\rm true}
\end{align}
where $R(p, E| p_{\rm true}, E_{\rm true}$ is the instrument response and $t_{\rm obs}$ is the observation time

@@

\begin{align}
   R(p, E|p_{\rm true}, E_{\rm true}) = &A_{\rm eff}(p_{\rm true}, E_{\rm true}) \times\\
    &PSF(p|p_{\rm true}, E_{\rm true}) \times\\
    &E_{\rm disp}(E|p_{\rm true}, E_{\rm true})
\end{align}
where:
\begin{itemize}
\item $A_{\rm eff}(p_{\rm true}, E_{\rm true})$ is the effective collection area of the detector. It is the product
  of the detector collection area times its detection efficiency at true energy $E_{\rm true}$ and position $p_{\rm true}$.
\item $PSF(p|p_{\rm true}, E_{\rm true})$ is the point spread function. It gives the probability of
  measuring a direction $p$ when the true direction is $p_{\rm true}$ and the true energy is $E_{\rm true}$.
  \gammaray instruments consider the probability density of the angular separation between true and reconstructed directions
  $\delta p = p_{\rm true} - p$, i.e. $PSF(\delta p|p_{\rm true}, E_{\rm true})$.
\item $E_{\rm disp}(E|p_{\rm true}, E_{\rm true})$ is the energy dispersion. It gives the probability to
  reconstruct the photon at energy $E$ when the true energy is $E_{\rm true}$ and the true position $p_{\rm true}$.
  \gammaray instruments consider the probability density of the migration $\mu=\frac{E}{E_{\rm true}}$,
  i.e. $E_{\rm disp}(\mu|p_{\rm true}, E_{\rm true})$.
\end{itemize}

\gammaray data at the Data Level 3 (DL3) therefore consist of lists of gamma-like events and their
corresponding instrument response functions (IRFs). The latter include the effective area ($A_{\rm eff}$),
point spread function (PSF) and energy dispersion ($E_{\rm disp}$).

An additional component of DL3 IRFs is the residual hadronic background model $Bkg$.
It represents the intensity of charged particles misidentified as gamma-rays that are expected
during an observation. It is defined as a function of the measured position in the field-of-view
and measured energy.

In total, the expected number of events in a \gammaray observation is given by:
\begin{align}
  N(p, E, \hat{\theta})\ {\rm d}p\ {\rm d}E =  &E_{\rm disp} \star \left[ PSF \star \left( A_{\rm eff} \times t_{\rm obs} \times \Phi(\hat{\theta}) \right) \right]\\
                       & + Bkg(p, E)
\end{align}

Finally, predicted and observed events, $N_{obs}$, can be then combined in a likelihood function,
$\mathcal{L}(\hat{\theta}, N_{obs})$, usually Poissonian, that is maximised to obtain the best-fit parameters of the flux model, $\bar{\theta}$.

@@

After all data has been projected into the same geometry, it is typically
required to improve the residual hadronic background estimate. As residual hadronic
background models can be subject to significant systematic uncertainties,
the background estimated can be improved by taking into account actual data
from regions without known \gammaray sources. This includes methods 
such as the ring or the field-of-view background techniques or
background measurements performed within, e.g. reflected regions~\citep{Berge07}.
Data measured at the field-of-view or energy boundaries of the instrument is typically
associated with a systematic uncertainty in the IRF. For this reason this part 
of the data is often excluded from subsequent analysis by defining regions of
 "safe" data in the spatial as well as energy dimension.
All of these data reduction steps are performed by classes and functions
implemented in the \code{gammapy.makers} subpackage (see Section~\ref{ssec:gammapy-makers}).

The counts data and the reduced IRFs in the form of maps are bundled into "datasets"
that represent the fourth data level (DL4). These reduced datasets can be written to disk,
in a format specific to \gammapy to allow users to read them back at any time later
for modeling and fitting. Different variations of such datasets support different 
analysis methods and fit statistics. The datasets are used to perform joint-likelihood
fitting allowing to combine different measurements, e.g. from different observations
but also from different instruments or event classes. They can also be used for binned
simulation as well as event sampling to simulate DL3 events data.
The various DL4 objects and the associated functionalities are
implemented in the \code{gammapy.datasets} subpackage (see Section~\ref{ssec:gammapy-datasets}).

The next step is then typically to model and fit the datasets, either
individually, or in a joint likelihood analysis. For this purpose \gammapy
provides a uniform interface to multiple fitting backends. In addition to
providing a variety of built in models, including spectral,
spatial and temporal model classes to describe the \gammaray emission in the sky,
custom user-defined models are also supported.
Spectral models can be simple analytical models or more complex ones from radiation
mechanisms of accelerated particle populations (e.g. inverse Compton or $\pi^{o}$ decay).
Independently or subsequently to the global modelling, the data can be
re-grouped to compute flux points, light curves and flux as well as significance
maps in energy bands.
The modelling and fitting functionalities are implemented in the \code{gammapy.modeling},
\code{gammapy.estimators} and \code{gammapy.stats} subpackages (see respectively
Section~\ref{ssec:gammapy-modeling}, \ref{ssec:gammapy-estimators} and \ref{ssec:gammapy-stats}).

\section{\gammapy Package}
\label{sec:gammapy-package}
\subsection{Overview}
\label{ssec:overview}
%
%
The \gammapy package is structured into multiple sub-packages. The definition
of the content of the different sub-packages follows mostly the stages in the
data reduction workflow described in the previous section. Sub-packages
either contain data structures representing data at different data reduction
levels or contain algorithms to transition between the different data reduction
levels.

>>>>

\newcommand{\hess}{H.E.S.S.\xspace}
\newcommand{\hawc}{HAWC\xspace}
\newcommand{\veritas}{VERITAS\xspace}
\newcommand{\magic}{MAGIC\xspace}
\newcommand{\astri}{ASTRI\xspace}
\newcommand{\iact}{IACT\xspace}
\newcommand{\iacts}{IACTs\xspace}
\newcommand{\cta}{CTA\xspace}
\newcommand{\swgo}{SWGO\xspace}
\newcommand{\irf}{IRF\xspace}
\newcommand{\irfs}{IRFs\xspace}
\newcommand{\fermi}{\textit{Fermi}-LAT\xspace}
\newcommand{\gammaray}{$\gamma$-ray\xspace}
\newcommand{\gammarays}{$\gamma$ rays\xspace}
\newcommand{\gadf}{GADF\xspace}
\newcommand{\milagro}{MILAGRO\xspace}
\newcommand{\github}{GitHub\xspace}

@@

\abstract
	{
		Traditionally, \gammaray astronomy has been conducted
		by small collaborations of scientists employing proprietary data and analysis software.
		However, the next generation of \gammaray instruments,
		such as the the Cherenkov Telescope Array (CTA), will be operated as open observatories.
		Alongside the data, the observatories will make available to the astronomical community
		software tools for their analysis. This necessity prompted the development, in the last decade,
		of open high-level astronomy software customised for high energy astrophysics.
	}
	{
		In this article, we present \gammapy, an open-source python package for the analysis of \gammaray data,
		and illustrate the functionalities of the first long-term release of the software, version 1.0.
		Built on the modern python scientific ecosystem, \gammapy provides a uniform platform for reducing and
		modelling data from different gamma-ray instruments. \gammapy complies with several well-established
		data conventions in high-energy astrophysics, providing serialised data products that are interoperable
		with other software.
	}
	{
		Starting from event list and instrument response functions,
		\gammapy provides pipelines for creating reduced data binned in energy and coordinates.
		To facilitate the rejection of the residual hadronic background, several techniques for background estimation
		are implemented in the package.
		After the data are binned, the flux and morphology of one or more \gammaray sources can be estimated
		using Poisson maximum likelihood fitting
		and assuming a variety of spectral, temporal, and spatial models.
		Estimation of flux points, likelihood profiles and light curves is also implemented.
	}
	{
		After describing the structure of the package, we demonstrate the capabilities of \gammapy
		performing both traditional and novel analyses in gamma-ray astronomy with publicly available data:
		spectral and spectro-morphological analyses, estimation of the light curve of a source.
		The flexibility and power of the software are displayed in a final multi-instrument example,
		where data sets from different instruments, at different stages of data reduction,
		are simultaneously fitted with a flux model produced by another astrophysical software.
	}{}

@@

%% \gammaray astronomy
%% CN: I think we should use the expression Water Cherenkov Detectors and use the
%% abbreviation (WCD) all along the text, later we use "particle samplers"
\gammaray astronomy is a rather young field of research. The \gammaray~
range of the electromagnetic spectrum provides us with insight into the
most energetic processes in the universe such as those accelerating particles in the surroundings of
black holes, or in remnants of supernova explosions. As in other
branches of astronomy, \gammarays can be observed by both
satellite as well as ground based instruments.
Ground-based instruments use the Earth's atmosphere as a particle detector.
Very-high-energy (VHE) cosmic \gammarays interact in the atmosphere and
create a large shower of secondary particles that can be observed from the ground.
Ground-based \gammaray astronomy relies on these extensive air showers to detect the
primary \gammaray photons and infer their direction and energy.
VHE \gammaray astronomy covers the energy range from fews tens of ${\rm GeV}$ up to the ${\rm PeV}$.
There are two main categories of ground-based instruments: 
Imaging Atmospheric Cherenkov Telescopes (IACT) and 
Water Cherenkov observatories. IACTs obtain images of the atmospheric showers
by detecting the Cherenkov radiation emitted by the cascading charged particles and
use these images to reconstruct the properties of the incident particle.
Those instruments have a limited field of view (FoV) and duty cycle, but
good energy and angular resolution.

\begin{figure*}[t]
	\centering
	\includegraphics[width=\textwidth]{figures/big-picture.pdf}
	\caption{
		Core idea and relation of \gammapy to different \gammaray instruments
		and the gamma astro data formats (GADF). The top left shows the
		group of current and future pointing instruments based on the 
		imaging atmospheric Cherenkov technique (IACT). This includes
		instruments such as the  Cherenkov Telescope Array (CTA),
		the High Energy Stereoscopic System (H.E.S.S.), the
		Major Atmospheric Gamma Imaging Cherenkov Telescopes (MAGIC),
		and the Very Energetic Radiation Imaging Telescope Array System (VERITAS).
		The lower left shows the group of all-sky instruments such as the
		Fermi Large Area Telescope (Fermi-LAT) amd the High Altitude 
		Water Cherenkov Observatory (HAWC). The calibrated data of all those
		instruments can be converted and stored into the common GADF data format.
		\gammapy can read data stored in the GADF format.
		The \gammapy package is not a part of any instrument, but instead
		provides a common interface to the data and analysis of all
		these \gammaray instruments. This way users can also easily combine data from
		different instruments and perform a joint analysis.
		\gammapy is built on the scientific python ecosystem, and the required dependencies
		are shown below the \gammapy logo.
	}
	\label{fig:big_picture}

@@

Water Cherenkov observatories detect particles directly from the tail of the
shower when it reaches the ground. These instruments have very
large field-of-view, large duty-cycle but higher energy threshold and
usually have lower signal to noise ratios compared to IACTs \citep{2015CRPhy..16..610D}.

%% Context
Ground-based \gammaray astronomy has been historically conducted
by experiments operated by independent collaborations, each relying
on their own proprietary data and analysis software developed as part of the
instrument. While this model has been successful so far, it does not
permit easy combination of data from several instruments and therefore,
limits the interoperability of existing facilities. This lack of
interoperability currently limits the full exploitation of the
available \gammaray data, especially because the different instruments often have
complementary sky coverages, and the various detection
techniques have complementary properties in terms of the energy range covered,
duty cycle and spatial resolution.

%TODO: better transition into context
The Cherenkov Telescope Array (CTA) will be the first ground-based \gammaray instrument to be operated
as an open observatory. Its high-level data (e.g. the event list) will be shared publicly after
some proprietary period, and the software required to analyze it will be distributed
as well. To allow the re-usability of existing instruments and their interoperability
requires open data formats and open tools that can support the various analysis methods
commonly used in the field.

%% Context : formats
%% CN: particle samplers -> WCD
In practice, the data reduction workflow of all \gammaray observatories
is remarkably similar. After data calibration, shower events are reconstructed and
gamma/hadron separation is applied to build lists of \gammaray-like events.
The lists of \gammaray events are then used to derive scientific results, such as spectra, sky maps
or light curves, taking into account the observatory's specific instrument response functions (IRF).
Once the data is reduced to a list of events, the information is independent of
the data-reduction process, and, eventually, of the detection technique. This implies,
for example, that hihg-level data from IACT and particle samplers can be represented
with the same data model. The efforts to prototype a format usable by various instruments
converged in the so-called \textit{Data Format for \gammaray Astronomy}
initiative~\citep{gadf_proc, gadf_universe}, abbreviated in
\texttt{gamma-astro-data-format} (\gadf). This proposes prototypical
specifications to produce files based on the flexible image transport system
(FITS) format~\citep{fits} encapsulating this high-level information. This is
realized by storing a list of gamma-like events with their measured quantities
(energy, direction, arrival time) and a parametrisation of the response of the
system. %% where? in the data units of a FITS file.

%% Context: develoment of open software
Python has become extremely popular as a scientific programming language,
in particular, in the field of data sciences. The success is
mostly attributed to the simple and easy to learn syntax, the ability to act as
a 'glue' language between different programming languages and last but not least
the rich ecosystem of packages and its open and supportive community, from
core numerical analysis packages such as \numpy~\citep{numpy} and
\scipy~\citep{2020SciPy-NMeth}, visualization libraries, e.g. Matplotlib \citep{matplotlib},
to optimization packages such as \iminuit \citep{iminuit}.

The \astropy~\citep{astropy} project was created in 2012 to build a community-developed
Python package for astronomical research. Among other things, it offers functionalities for
transforming and representing astronomical coordinates, manipulating physical quantities
and their units as well as reading and writing FITS files.

%% \gammapy: concept and goals
%%% Question: starting date? Is is relevant, what is the real start?
The \gammapy project started in 2015 with the objective of building a flexible and
efficient software library for very high-energy \gammaray data analysis \citep{gammapy_2015}.
This is possible thanks to the definition of a standardised common data format such as the \gadf, and thanks to
an open development with a broad community of contributors. The relation of \gammapy to the
different instruments and the \gadf data format is illustrated in Figure~\ref{fig:big_picture}.

@@

%% \gammapy: from first developments to validation and selection
%% CN: what about mentioning the joint Crab? or is it mentioned later?
The \hess collaboration released a limited test dataset (about 50 hours of
observations taken between 2004 and 2008) based  on the \gadf DL3 format \citep{HESS_DR1}.
This data release served as a basis for validation of open analysis tools,
including \gammapy \cite[see e.g.]{Mohrmann2019}.

%% Paper outline
In this article, we describe the general structure of the \gammapy package,
its main concepts and organisational structure. We start in
Section~\ref{sec:gammaray-data-analysis} with a general overview
of the data analysis workflow in very high-energy \gammaray astronomy. Then we
show how this workflow is reflected in the structure of the \gammapy package 
in Section~\ref{sec:gammapy-package}, while also
describing the various subpackages it contains. Section~\ref{sec:applications}
presents a number of applications, while Section~\ref{sec:gammapy-project}
finally discusses the project organization.

\section{Gamma-ray Data Analysis}
% CN: gamma-like here, was \gammaray-like in the previous section
\begin{figure*}[h!]
	\centering
	\includegraphics[width=1.\textwidth]{figures/data_flow.pdf}
	\caption{
		\gammapy sub-package structure and data analysis workflow. The top row
        defines the different levels of data reduction, from lists of gamma-like
        events on the left (DL3), to high-level scientific products products
        (DL5) on the right. The direction of the data flow is illustrated by the
        grey arrows. The gray folder icons represent the different sub-packages
        in \gammapy and their names. Below each icon there is a list of the most
        important objects defined in the sub-package.
    }
	\label{fig:data_flow}
\end{figure*}
%
\label{sec:gammaray-data-analysis}
% This might not be the bast place compared to introduction
The data reduction process in \gammaray astronomy is usually split into two parts.
The first one deals with the data processing from camera measurement, calibration, event
reconstruction and selection to yield a list of reconstructed \gammaray event candidates.
This sequence, sometimes referred to as low-level analysis, is usually very specific to
a given observation technique and even to a given instrument.

The other sequence, referred to as high-level analysis, deals with the extraction of physical
quantities related to \gammaray sources and the production of high-level products such as spectra,
light curves and catalogs. The methods applied here are more generic and are broadly
shared across the field. They also frequently imply joint analysis of multi-instrument data.
%% CN: I don't understand this: since the high-level methods are the same, then they "imply"
%% joint analyses?? Maybe they eventually allow or permit?

@@

To extract physically relevant information, such as the flux, spatial or spectral shape of one or more sources,
an analytical model is commonly adopted to describe the intensity of gamma-ray sources as a function of the energy,
$E_{\rm true}$, and of the position in the field of view, $p_{\rm true}$:
$$ \Phi(p_{\rm true}, E_{\rm true}, \hat{\theta}) \quad \mathrm{[TeV^{-1}cm^{-2}s^{-1}]} $$
where $\hat{\theta}$ is a set of model parameters that can be adjusted in a fit. To convert this analytical flux model
into a prediction on the number of gamma-ray events, $N_{\rm pred}$, with their estimated energy $E$ and position $p$, the model
is convolved through the response function of the instrument.

%% CN: not sure if we already commented about this, but align is artificially
%% numbering every new line as a new equation
%% I think
%% \begin{equation}
%%   \begin{split}
%%     ...
%%   \begin{split}
%% \begin{equation}
%% will avoid to number each new line

In the most general way, we can write the expected number of detected events from the sky model $\Phi$ at measured position $p$ and energy $E$, for
a given set of parameters $\hat{\theta}$, as
\begin{align}
   N(p, E, \hat{\theta})\,{\rm d}p\,{\rm d}E = &t_{\rm obs} \int_{E_{\rm true}} \int_{p_{\rm true}}  R(p, E|p_{\rm true}, E_{\rm true})\\
   &\times \Phi(p_{\rm true}, E_{\rm true}, \hat{\theta} ) {\rm d}E_{\rm true} {\rm d}p_{\rm true}
\end{align}
where $R(p, E| p_{\rm true}, E_{\rm true})$ is the instrument response and $t_{\rm obs}$ is the observation time

@@

\begin{align}
   R(p, E|p_{\rm true}, E_{\rm true}) = &A_{\rm eff}(p_{\rm true}, E_{\rm true}) \times\\
    &PSF(p|p_{\rm true}, E_{\rm true}) \times\\
    &E_{\rm disp}(E|p_{\rm true}, E_{\rm true})
\end{align}
where:
\begin{itemize}
\item $A_{\rm eff}(p_{\rm true}, E_{\rm true})$ is the effective collection area of the detector. It is the product
  of the detector collection area times its detection efficiency at true energy $E_{\rm true}$ and position $p_{\rm true}$.
\item $PSF(p|p_{\rm true}, E_{\rm true})$ is the point spread function (PSF). It gives the probability of
  measuring a direction $p$ when the true direction is $p_{\rm true}$ and the true energy is $E_{\rm true}$.
  \gammaray instruments consider the probability density of the angular separation between true and reconstructed directions
  $\delta p = p_{\rm true} - p$, i.e. $PSF(\delta p|p_{\rm true}, E_{\rm true})$.
\item $E_{\rm disp}(E|p_{\rm true}, E_{\rm true})$ is the energy dispersion. It gives the probability to
  reconstruct the photon at energy $E$ when the true energy is $E_{\rm true}$ and the true position $p_{\rm true}$.
  \gammaray instruments consider the probability density of the migration $\mu=\frac{E}{E_{\rm true}}$,
  i.e. $E_{\rm disp}(\mu|p_{\rm true}, E_{\rm true})$.
\end{itemize}

%% CN: except for the figures we never introduced what the DL3 or the Data Level acronym is
%% maybe add (see Fig.~1)
\gammaray data at the Data Level 3 (DL3) therefore consist of lists of gamma-like events and their
corresponding instrument response functions. The latter include the effective area ($A_{\rm eff}$),
point spread function and energy dispersion ($E_{\rm disp}$).
%% CN: this paragraph repeats what is commented in the previous formula and itemize and can be removed

An additional component of DL3 IRFs is the residual hadronic background model $Bkg$.
It represents the intensity of charged particles misidentified as \gammarays that are expected
during an observation. It is defined as a function of the measured position in the field-of-view
and measured energy.

In total, the expected number of events in a \gammaray observation is given by:
\begin{align}
  N(p, E, \hat{\theta})\ {\rm d}p\ {\rm d}E =  &E_{\rm disp} \star \left[ PSF \star \left( A_{\rm eff} \times t_{\rm obs} \times \Phi(\hat{\theta}) \right) \right]\\
                       & + Bkg(p, E)
\end{align}
%% CN: Instead of rewriting this formula and introducing this new notation for the convolution,
%% can't we just say that, when predicting counts, $Bkg(p, E)$ has to be summed to the convolution
%% (or integral) in the previous equation?

Finally, predicted and observed events, $N_{obs}$, can then be combined in a likelihood function,
$\mathcal{L}(\hat{\theta}, N_{obs})$, usually Poissonian, that is maximised to obtain the best-fit parameters of the flux model, $\hat{\theta}$.

@@

%% CN: I would actually distingusih two cases here:
%% 1) the Bkg is already provided as a parametrisation
%% 2) the Bkg is not provided and, thus, has to be estimated from the obs.
After all data has been projected into the same geometry, it is typically
required to improve the residual hadronic background estimate. As residual hadronic
background models can be subject to significant systematic uncertainties,
the background estimated can be improved by taking into account actual data
from regions without known \gammaray sources. This includes methods 
such as the ring or the field-of-view background techniques or
background measurements performed within, e.g. reflected regions~\citep{Berge07}.
Data measured at the field-of-view or energy boundaries of the instrument is typically
associated with a systematic uncertainty in the IRF. For this reason this part 
of the data is often excluded from subsequent analysis by defining regions of
 "safe" data in the spatial as well as energy dimension.
All of these data reduction steps are performed by classes and functions
implemented in the \code{gammapy.makers} subpackage (see Section~\ref{ssec:gammapy-makers}).

%% CN: for 1D analysis datasets, DL4 can actually be written as OGIP files,
%% so as reduced datasets in a general format, not specific to Gammapy.
%% Maybe you can briefly mention that this is due to the fact that the 1D Poisson
%% spectral analysis is liong consilidated, but this also means that Gammapy 's
%% dataset can handle lower energy data, e.g. X-ray.
%% We also said in the abstract that Gammapy is compliant with other specifications
%% adopted in high-energy astrophysics.
The counts data and the reduced IRFs in the form of maps are bundled into "datasets"
that represent the fourth data level (DL4). These reduced datasets can be written to disk,
in a format specific to \gammapy to allow users to read them back at any time later
for modeling and fitting. Different variations of such datasets support different 
analysis methods and fit statistics. The datasets can be used to perform a joint-likelihood
fit, allowing one to combine different measurements, e.g. from different observations
but also from different instruments or event classes. They can also be used for binned
simulation as well as event sampling to simulate DL3 events data.
The various DL4 objects and the associated functionalities are
implemented in the \code{gammapy.datasets} subpackage (see Section~\ref{ssec:gammapy-datasets}).

The next step is then typically to model and fit the datasets, either
individually, or in a joint likelihood analysis. For this purpose \gammapy
provides a uniform interface to multiple fitting backends. In addition to
providing a variety of built-in models, including spectral,
spatial and temporal model classes to describe the \gammaray emission in the sky,
custom user-defined models are also supported.
Spectral models can be simple analytical models or more complex ones from radiation
mechanisms of accelerated particle populations (e.g. inverse Compton or $\pi^{o}$ decay).
Independently or subsequently to the global modelling, the data can be
re-grouped to compute flux points, light curves and flux as well as significance
maps in different energy bands.
The modelling and fitting functionalities are implemented in the \code{gammapy.modeling},
\code{gammapy.estimators} and \code{gammapy.stats} subpackages (see respectively
Section~\ref{ssec:gammapy-modeling}, \ref{ssec:gammapy-estimators} and \ref{ssec:gammapy-stats}).

\section{\gammapy Package}
\label{sec:gammapy-package}
\subsection{Overview}
\label{ssec:overview}
%
%
The \gammapy package is structured into multiple sub-packages. The definition
of the content of the different sub-packages follows mostly the stages of the
data reduction workflow described in the previous section. Sub-packages
either contain structures representing data at different reduction
levels or algorithms to transition between these different levels.
